---
title: "coupledAbmOdeAnalysis"
author: "Inan Bostanci & Jakob Rehmann"
date: "2025-08-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(patchwork)
library(zoo) 
library(sf)
library(tmap)
library(tmaptools)
library(scales)
library(gganimate)

source("masterJR-utils.R")
```

Some Defaults and chosen values

```{r}
POPULATION_BERLIN <- 3782202
POPULATION_BRANDENBURG <- 2556747

plot_start_date <- ymd("2020-03-01")
# plot_end_date <- ymd("2020-12-31")
plot_end_date <- ymd("2021-03-01")

brand_districts <- c("Dahme-Spreewald", "Potsdam", "Barnim", "Märkisch-Oderland", "Oberhavel", "Havelland", "Teltow-Fläming", "Brandenburg an der Havel", "Cottbus", "Spree-Neiße", "Frankfurt (Oder)", "Potsdam-Mittelmark", "Elbe-Elster", "Oder-Spree", "Oberspreewald-Lausitz", "Ostprignitz-Ruppin", "Prignitz", "Uckermark")
```

# RKI data on reported cases

RKI provides the daily number of new confirmed COVID-19 cases in Germany, available at the county level (IdLandkreis, starts with "11" for Berlin, "12" for Brandenburg), broken down by age group and gender. The data can be differentiated into "Meldedatum", which is the date of reporting, and "Refdatum", which is either the assumed date of the actual onset, or if unknown, the date of reporting.

The data is available in CSV format and can be downloaded from <https://github.com/robert-koch-institut/SARS-CoV-2-Infektionen_in_Deutschland_Archiv>

For the ODE fit, it seems that we used "Refdatum", which is the date of the actual onset or reporting date, and not "Meldedatum", which is the date of reporting. However, for the ABM, I think we used "Meldedatum" - this might make sense if the ABM enables to simulate the reporting process, while the ODE's I-compartment simulates the actual disease progression.

```{r}
# Helper function to process and aggregate RKI data
process_rki_data <- function(df, date_col, region_prefix = "11") {
  df %>%
    filter(str_starts(as.character(IdLandkreis), region_prefix)) %>%
    group_by(date = as_date(.data[[date_col]])) %>%
    summarise(daily_new = sum(AnzahlFall, na.rm = TRUE), .groups = "drop") %>%
    complete(date = seq(min(date), max(date), by = "1 day"), fill = list(daily_new = 0)) %>%
    arrange(date)
}
```

Load RKI data from Github

```{r}
# Load data from the RKI GitHub repository
rki_repo_url <- "https://github.com/robert-koch-institut/SARS-CoV-2-Infektionen_in_Deutschland_Archiv/raw/refs/heads/main/Archiv/2021-05-31_Deutschland_SarsCov2_Infektionen.csv.xz"

# Download to a temp file
tmpfile <- tempfile(fileext = ".csv.xz")
download.file(rki_repo_url, tmpfile, mode = "wb")

# Read from compressed file directly
raw_data <- read_csv(tmpfile)
```

calculate incidence for three study areas

Note: when 'align = "right"' rollmean centers the mean around the right date of the range (for 7 observations between Mon & Sun, the value would be set to Sunday)

```{r}
# Load data and process both Refdatum 

date_type <- "Refdatum"
rki_berlin <- process_rki_data(raw_data, date_type) %>% 
  mutate(incidence = rollsum(daily_new, 7, fill = NA,align = "right") / POPULATION_BERLIN * 1e5)

rki_brand <- process_rki_data(raw_data, date_type, region_prefix = "12") %>%
  mutate(incidence = rollsum(daily_new, 7, fill = NA,align = "right") /  POPULATION_BRANDENBURG * 1e5)

rki_combined<- rbind(rki_brand, rki_berlin) %>% group_by(date) %>% 
  summarise(daily_new = sum(daily_new))%>%
  mutate(incidence = rollsum(daily_new, 7, fill = NA,align = "right") / (POPULATION_BRANDENBURG + POPULATION_BERLIN) * 1e5)


```
plausibility checks

It seems that inan used the ReferenzDatum. So I will use this for now as well
```{r}
tmpfile <- tempfile(fileext = ".csv")
download.file("https://raw.githubusercontent.com/robert-koch-institut/COVID-19_7-Tage-Inzidenz_in_Deutschland/refs/heads/main/COVID-19-Faelle_7-Tage-Inzidenz_Bundeslaender.csv", tmpfile, mode = "wb")

# Read from compressed file direct
rki2  <- read_csv(tmpfile)

# Plausibility of Brandenburg

rki2_brand <- rki2 %>% filter(Bundesland_id == "12") %>% filter(Altersgruppe == "00+")

ode_incidence_brandenburg <- read_csv("/Users/jakob/git/shared-svn/projects/episim/matsim-files/snz/BerlinV2/episim-input/ode_br_infectious_250212.csv") %>%
  mutate(new_infections = infectious / 3 * POPULATION_BRANDENBURG  ) %>% 
  mutate(incidence = rollsum(new_infections, 7, fill = NA, align = "right")/ POPULATION_BRANDENBURG * 100000)


ggplot(rki_brand %>% filter(wday(date) ==5)) + geom_point(aes(date,incidence))  + 
  scale_x_date(date_breaks = "1 month",date_labels = "%b\n'%y")  +
  geom_line(data = rki2_brand %>% filter(Meldedatum < ymd("2020-12-31")), aes(Meldedatum,`Inzidenz_7-Tage`))+
  geom_line(data = ode_incidence_brandenburg, aes(time, incidence), col = "yellow")


# Plausibility of Berlin

rki2_berlin <- rki2 %>% filter(Bundesland_id == "11") %>% filter(Altersgruppe == "00+")

ode_incidence_berlin <- read_csv("/Users/jakob/git/shared-svn/projects/episim/matsim-files/snz/Brandenburg/episim-input/ode_be_infectious_250211.csv") %>%
  mutate(new_infections = infectious / 3 * POPULATION_BRANDENBURG  ) %>% 
  mutate(incidence = rollsum(new_infections, 7, fill = NA, align = "right")/ POPULATION_BRANDENBURG * 100000)


ggplot(rki_berlin %>% filter(wday(date) ==5)) + geom_point(aes(date,incidence))  + 
  scale_x_date(date_breaks = "1 month",date_labels = "%b\n'%y")  +
  geom_line(data = rki2_berlin %>% filter(Meldedatum < ymd("2020-12-31")), aes(Meldedatum,`Inzidenz_7-Tage`))+
  geom_line(data = ode_incidence_berlin, aes(time, incidence), col = "yellow")
```

# Episim Data

Setup

```{r}

image_output_path <- "/Users/jakob/Dropbox/Documents/04_Work-VSP/007_papers/2025-08-14-abm-coupling/images/"

common_style <- list(
  scale_color_manual(values = c("RKI" = "darkorange","EpiSim" = "blue")),
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b\n'%y"
  ),
  scale_y_log10(limits = c(0.1,1000), labels = label_number()),
  theme_minimal(base_size = 14),
  theme(legend.position = "bottom"),
  labs(x = "Date", y = "SARS-CoV-2 Incidence", col = NULL)
)

```



## Combined: Berlin + Brandenburg
```{r}

combined_directory <- "/Users/jakob/git/public-svn/matsim/scenarios/countries/de/episim/battery/jakob/2026-02-03/1-both-br/"


combined_districts <- c("Berlin", brand_districts)

# combined_info <- read_delim(paste0(combined_directory, "_info.txt"), delim = ";") %>%
  # filter(thetaFactor == 0.7,pHouseholds == 0.0, workLeisureAdjustment == FALSE,importToBerlin == FALSE) 

combined_info <- read_delim(paste0(combined_directory, "_info.txt"), delim = ";") %>% 
    filter(thetaFactor %in% c(0.75), maxOutdoorFraction == 1.0, importMultSpring %in% c(0.25),importToBerlin == FALSE)

# %>% 
  # filter(pHouseholds == 0.35, importMult == 2)

# combined_info %>% view()

combined_raw <- read_combine_episim_output_zipped(combined_directory, combined_info,"infections.txt.csv", c("date","nShowingSymptomsCumulative","nSusceptible","district")) 



combined_params <- get_run_parameters(combined_directory)

combined_filtered_berlin <- combined_raw  %>% filter(district == "Berlin")

combined_filtered_brand <- combined_raw %>% 
  filter(district %in% brand_districts) %>% 
  group_by(across(all_of(c("date",combined_params)))) %>%
  summarise(nShowingSymptomsCumulative = sum(nShowingSymptomsCumulative), nSusceptible = sum(nSusceptible)) %>% 
  mutate(district = "Brandenburg")

combined_filtered_combined <- combined_raw %>% 
  filter(district %in% combined_districts) %>% 
  group_by(across(all_of(c("date",combined_params)))) %>%
  summarise(nShowingSymptomsCumulative = sum(nShowingSymptomsCumulative), nSusceptible = sum(nSusceptible)) %>% 
  mutate(district = "Combined")

combined_filtered <- rbind(combined_filtered_berlin, combined_filtered_brand, combined_filtered_combined)


combined_incidence <- convert_infections_into_incidence(combined_directory,combined_filtered,FALSE)


combined_incidence_filtered <- combined_incidence %>% 
  select(date, incidence, district) %>% 
  mutate(scenario = "EpiSim") %>% 
  rbind(rki_combined %>% select(date, incidence = incidence) %>% mutate(scenario = "RKI",district = "Combined")) %>% 
  rbind(rki_berlin %>% select(date, incidence = incidence) %>% mutate(scenario = "RKI",district = "Berlin")) %>% 
  rbind(rki_brand %>% select(date, incidence = incidence) %>% mutate(scenario = "RKI",district = "Brandenburg")) %>% 
  filter(date >= plot_start_date, date <= plot_end_date) %>% 
  filter(lubridate::wday(date, label = TRUE)=="Thu")

# ggplot(combined_incidence_filtered %>% filter(scenario !="RKI")) + geom_point(aes(date,incidence, col = district))
  

combined_plot_combined <- ggplot() +
  geom_point(data = combined_incidence_filtered %>% filter(district == "Combined"),mapping = aes(date,incidence, col = scenario)) + 
  labs(title ="Incidence in Berlin + Brandenburg", subtitle = "Berlin + Brandenburg modelled as ABM") +
  common_style

combined_plot_combined

combined_plot_berlin <- ggplot() +
  geom_point(data = combined_incidence_filtered %>% filter(district == "Berlin"),mapping = aes(date,incidence, col = scenario),show.legend = FALSE) + 
  labs(title ="Incidence in Berlin only", subtitle = "Berlin + Brandenburg modelled as ABM") +
  common_style

combined_plot_berlin

combined_plot_brand <- ggplot() +
  geom_point(data = combined_incidence_filtered %>% filter(district == "Brandenburg"), mapping = aes(date,incidence, col = scenario), show.legend = FALSE) + 
  labs(title ="Incidence in Brandenburg only", subtitle = "Berlin + Brandenburg modelled as ABM") +
  common_style

combined_plot_brand


# plot all three
combined_plot <- combined_plot_combined / (combined_plot_brand | combined_plot_berlin)  + plot_annotation(tag_levels = 'a')
# combined_plot

ggsave(paste0(image_output_path,"results_combined.pdf"), plot = combined_plot, width = 12, height = 8)


# ggplot(rki_berlin) + geom_point(aes(date,incidence), col = "red")

```


## Brandenburg

what happens if I change the max outdoor fraction for this simulation also to 0.8, then there will be more infections in the summer! 


```{r}

# ggplot(bind_rows(rki_berlin %>% mutate(area = "berlin"),rki_brand %>% mutate(area = "brand")) %>% filter(date <= plot_end_date)) + geom_point(aes(date,incidence, col = area))
 
# brand_directory <- "/Users/jakob/Dropbox/Documents/04_Work-VSP/007_papers/2025-08-14-abm-coupling/Data4Paper/2025-10-17/2-brand-quick/"

brand_directory <- "../public-svn/matsim/scenarios/countries/de/episim/battery/jakob/2026-02-03/2-brand-br/"


  # filter(thetaFactor %in% c(0.75),  workLeisureAdjustment == TRUE, importMultSpring == 0.25,importToBerlin == TRUE)

brand_info <- read_delim(paste0(brand_directory, "_info.txt"), delim = ";") %>% 
    filter(ode %in% c(1.0), importMult %in% c("x0.1"), maxOutdoorFraction == 1.0,thetaFactor == 0.75,workLeisureAdjustment == TRUE) #ode %in% c(1.0,4.0) #,"x0.075","x0.1","x0.125"


brand_raw <- read_combine_episim_output_zipped(brand_directory,brand_info,"infections.txt.csv", c("date","nShowingSymptomsCumulative","nSusceptible","district"))

brand_params <- get_run_parameters(brand_directory)

brand_filtered <- brand_raw %>% 
  filter(district %in% brand_districts) %>% 
  group_by(across(all_of(c("date",brand_params)))) %>%
  summarise(nShowingSymptomsCumulative = sum(nShowingSymptomsCumulative), nSusceptible = sum(nSusceptible)) %>% 
  mutate(district = "Brandenburg")

brand_incidence <- convert_infections_into_incidence(brand_directory,brand_filtered,FALSE)

brand_incidence_filtered <- brand_incidence %>% 
  select(date, incidence) %>% 
  mutate(scenario = "EpiSim") %>% 
  rbind(rki_brand %>% select(date, incidence) %>% mutate(scenario = "RKI")) %>% 
  filter(date >= plot_start_date, date <= plot_end_date) %>% 
  filter(lubridate::wday(date, label = TRUE)=="Thu")

brand_plot <- ggplot() +
  geom_point(data = brand_incidence_filtered, mapping = aes(date,incidence, col = scenario)) +
  labs(title ="Incidence in Brandenburg", subtitle = "Coupled Model: Brandenburg as ABM | Berlin as ODE") + 
  common_style

brand_plot
ggsave(paste0(image_output_path,"results_brand.pdf"), plot = brand_plot, width = 6, height = 4)

```

## Berlin

```{r}

berlin_directory <- "/Users/jakob/git/public-svn/matsim/scenarios/countries/de/episim/battery/jakob/2026-02-03/3-berlin/"

berlin_info <- read_delim(paste0(berlin_directory, "_info.txt"), delim = ";") %>% 
    filter(ode %in% c(1.0), importMult %in% c("x0.1"), maxOutdoorFraction == 1.0, thetaFactor == 0.75,workLeisureAdjustment == TRUE) #ode %in% c(1.0,4.0) #,"x0.075","x0.1","x0.125"
 # filter(ode == 1., thetaFactor %in% c(0.7,1.0))


berlin_raw <- read_combine_episim_output_zipped(berlin_directory,berlin_info,"infections.txt.csv", c("date","nShowingSymptomsCumulative","nSusceptible","district"))

berlin_filtered <- berlin_raw %>% filter(district == "Berlin")

berlin_incidence <- convert_infections_into_incidence(berlin_directory,berlin_filtered,FALSE)

berlin_incidence_filtered <-  berlin_incidence %>% 
  # mutate(scenario = paste0("EpiSim:",thetaFactor)) %>% 
    mutate(scenario = "EpiSim") %>%
  select(date, incidence, scenario) %>% 
  rbind(rki_berlin %>% select(date, incidence) %>% mutate(scenario = "RKI")) %>% 
  filter(date >= plot_start_date, date <= plot_end_date) %>% 
  filter(lubridate::wday(date, label = TRUE)=="Thu")

berlin_plot <- ggplot() +
  geom_point(data = berlin_incidence_filtered,mapping = aes(date,incidence, col = scenario)) + 
  labs(title ="Incidence in Berlin", subtitle = "Coupled Model: Berlin as ABM | Brandenburg as ODE") +
  common_style 
    # scale_color_manual(values = c("RKI" = "red","EpiSim:1" = "grey", setNames("black", paste0("EpiSim:", THETA_FACTOR)))) 

berlin_plot

ggsave(paste0(image_output_path,"results_berlin.pdf"), plot = berlin_plot, width = 6, height = 4)

```




# Geospatial & Temporal Analysis

```{r}

# brand_directory <- "/Users/jakob/git/public-svn/matsim/scenarios/countries/de/episim/battery/jakob/2026-02-03/2-brand-br-disagg/2-brand/"

brand_directory <- "/Users/jakob/Dropbox/Documents/04_Work-VSP/007_papers/2025-08-14-abm-coupling/Data4Paper/2-brand/"

berlin_brandenburg_envelope <- st_read("/Users/jakob/Dropbox/Documents/04_Work-VSP/SHAPE_FILES/berlinBrandenburg/berlinBrandenburg.shp") %>% st_transform(4236)

# plot infection map
info_df <- read_delim(paste0(brand_directory, "_info.txt"), delim = ";")


info_df %>% pull(seed) %>% unique() %>% as.character()
runId <- info_df %>% filter(seed == 7564655870752978944) %>% filter(ode %in% c(1.0), importMult %in% c("x0.1"), maxOutdoorFraction == 1.0,thetaFactor == 0.75,workLeisureAdjustment == TRUE) %>% pull(RunId)


geospatial <- read_delim(paste0(brand_directory,runId,".infectionLoc.csv")) %>%
  st_as_sf(coords = c("home_lon","home_lat"), crs = 4236) %>%
  st_filter(berlin_brandenburg_envelope) %>%   
  mutate(infection_type = factor(infection_type, levels = c("normal","import"),labels = c("ABM Infections","ODE Import"))) %>% 
  mutate(date = ymd("2020-03-20") + daysSinceStart)


tmap_mode("plot")

geospatial_plot_hist <- geospatial %>% 
  ggplot() +
  geom_histogram(aes(date, fill = infection_type), binwidth = 7)  +
  scale_x_date(
    date_breaks = "1 month",      # show every month
    date_labels = "%b\n'%y"         # e.g., Jan 2025
  ) +
  # scale_y_log10() + 
  labs(title ="New Infections in Brandenburg", subtitle = "Coupled Model: Brandenburg as ABM | Berlin as ODE", x = "Date", y = "New Infections") +
  theme_minimal(base_size = 14) 

geospatial_plot_hist

ggsave(paste0(image_output_path,"results_geospatial_hist.pdf"), plot = geospatial_plot_hist, width = 6, height = 4)

for(startDay in seq(0,360,by = 7)){
  endDay <- startDay + 7
  map <- tm_basemap("Esri.WorldGrayCanvas") + 
    tm_shape(geospatial %>% filter(daysSinceStart > startDay, daysSinceStart <= endDay), bbox = bb(berlin_brandenburg_envelope)) +
    tm_dots(fill = "infection_type", fill_alpha = 0.5,fill.scale = tm_scale(values = c("ODE Import" = "red", "ABM Infections" = "blue"))) +
    tm_title(paste0("Day ",startDay," - ",endDay))
  print(map)
}


geospatial_for_video <- geospatial %>% mutate(time = factor(date))

  # filter(daysSinceStart > startDay, daysSinceStart <= endDay)
tmap_for_video <- 
  tm_basemap("Esri.WorldGrayCanvas") +
  tm_shape(geospatial_for_video,bbox = bb(berlin_brandenburg_envelope)) +
  tm_dots(
    fill = "infection_type",
    fill_alpha = 0.5,
    fill.scale = tm_scale(values = c("ODE Import" = "red", "ABM Infections" = "blue")),
    fill.legend = tm_legend(title = "Infection source") 
    ) +
  tm_title("New Infections\nCoupled Model: Brandenburg as ABM | Berlin as ODE") +
  tm_layout(
    frame = FALSE,
    bg.color = "white",
    
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.8,
    legend.text.size = 0.9,
    legend.title.size = 1,
    
    title.size = 1.2
  ) +
  tm_options(facet.max = 365) + 
  tm_animate("time")
  
    
  
tmap_animation(tmap_for_video,filename = paste0(image_output_path, "map.mp4"))
  

```



## 2026-03-02


```{r}


shp_lk <- st_read("/Users/jakob/Dropbox/Documents/04_Work-VSP/SHAPE_FILES/landkreise-DE/landkreise-in-germany.shp") %>% filter(name_1 == "Brandenburg") %>% select(district = name_2)


flows <- read_csv("/Users/jakob/git/matsim-episim-vsp/flows.csv")
pop <- read_csv("/Users/jakob/git/matsim-episim-vsp/pop.csv")

rel_districts = c("Berlin",brand_districts)

flows_to_Berlin <- flows %>% filter(from %in% brand_districts & to == "Berlin") %>% select(district = from, commuters_to_berlin = cnt)
flows_from_Berlin <- flows %>% filter(from == "Berlin" & to %in% brand_districts) %>% select(district = to, commuters_from_berlin = cnt)
pop2 <- pop %>% filter(district %in% brand_districts) %>% rename(pop = cnt) 

flow_and_commuters <- pop2 %>%
  left_join(flows_to_Berlin, by = "district") %>%
  left_join(flows_from_Berlin, by = "district") %>% 
  mutate(commuters_total = commuters_to_berlin + commuters_from_berlin) %>% 
  left_join(shp_lk, by = "district")
  # filter(from!="Berlin")



```



Linking commuter flows to incidence
```{r}

brand_lk_filtered <- brand_raw %>% 
  filter(district %in% brand_districts)

brand_lk_incidence <- convert_infections_into_incidence(brand_directory,brand_lk_filtered,FALSE)

# first wave
brand_lk_incidence_max_1 <- brand_lk_incidence %>% 
  group_by(district) %>% 
  filter(date <= ymd("2020-06-01")) %>% 
  summarise(max_incidence_1 = max(incidence), max_date_1 = date[which.max(incidence)],total_cases_1 = sum(infections))  %>% 
  mutate(max_day_1 = as.numeric(max_date_1 - ymd("2020-02-25")))

# second wave
brand_lk_incidence_max_2 <- brand_lk_incidence %>% 
  group_by(district) %>% 
  filter(date > ymd("2020-06-01")) %>% 
  summarise(max_incidence_2 = max(incidence), max_date_2 = date[which.max(incidence)],total_cases_2 = sum(infections))  %>% 
  mutate(max_day_2 = as.numeric(max_date_2 - ymd("2020-02-25")))


# tmap_mode("plot")
# #cols4all::c4a_gui() to choose color scheme
# tm_basemap("Esri.WorldGrayCanvas") + 
#   tm_shape(brand_lk_incidence_max_1) + 
#   tm_polygons(fill = "max_incidence", fill.scale = tm_scale_continuous(values = "brewer.or_rd")) +
#   tm_title("Max Incidence")


# df_template <- data.frame(
#   district = c(
#     "Barnim",
#     "Brandenburg an der Havel",
#     "Cottbus",
#     "Dahme-Spreewald",
#     "Elbe-Elster",
#     "Frankfurt (Oder)",
#     "Havelland",
#     "Märkisch-Oderland",
#     "Oberhavel",
#     "Oberspreewald-Lausitz",
#     "Oder-Spree",
#     "Ostprignitz-Ruppin",
#     "Potsdam",
#     "Potsdam-Mittelmark",
#     "Prignitz",
#     "Spree-Neiße",
#     "Teltow-Fläming",
#     "Uckermark"
#   ),
#   commuters = c(
#     31195, 2535, 1738, 25317,934, 1896,
#     25023, 28179, 36866, 1018, 14381, 2770,
#     20307,22957, 805, 989 , 22235, 2582
#   )
# ) %>% inner_join(shp_lk) %>% st_as_sf()
# 
# tm_basemap("Esri.WorldGrayCanvas") + 
#   tm_shape(df_template) + 
#   tm_polygons(fill = "commuters") +
#   tm_title("# of Commuters")
# 


```

Merge
```{r}
merged <- flow_and_commuters %>%
  full_join(brand_lk_incidence_max_1, by =  "district") %>%
  full_join(brand_lk_incidence_max_2, by =  "district") %>% 
  mutate(pop = pop * 4, commuters_total = commuters_total *4) %>% 
  mutate(commuter_relative = commuters_total/pop) %>% st_as_sf()

areas <- merged %>% st_area() %>% units::drop_units()

# merged$area <- areas

merged <- merged %>% mutate(area_km2 = areas/1000/1000) %>% mutate(pop_density = pop/area_km2 )

# merged %>% view()

tmap_mode("plot")


cor1_pop <- tm_basemap("Esri.WorldGrayCanvas") + tm_shape(merged) + tm_polygons(fill = "pop")+ tm_title("Population")+   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )

cor2_pop_density <- tm_basemap("Esri.WorldGrayCanvas") + tm_shape(merged) + tm_polygons(fill = "pop_density")+ tm_title("Population Density")+   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )
cor3_commuters <- tm_basemap("Esri.WorldGrayCanvas") +tm_shape(merged) + tm_polygons(fill = "commuters_total") + tm_title("Commuters")+   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )
cor4_commuter_density <- tm_basemap("Esri.WorldGrayCanvas") + tm_shape(merged) + tm_polygons(fill = "commuter_relative") + tm_title("Commuters / Population")+   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )
cor5_max_incidence_wave1 <- tm_basemap("Esri.WorldGrayCanvas")+ tm_shape(merged) + tm_polygons(fill = "max_incidence_1") + tm_title("Max Incidence (1st Wave)")+   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )
cor6_max_incidence_wave2 <- tm_basemap("Esri.WorldGrayCanvas") + tm_shape(merged) + tm_polygons(fill = "max_incidence_2") + tm_title("Max Incidence (2nd Wave)") +   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )
cor7_max_date_wave1 <- tm_basemap("Esri.WorldGrayCanvas")+ tm_shape(merged) + tm_polygons(fill = "max_day_1") + tm_title("Date of Max Incidence (1st Wave)")+   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )
cor8_max_date_wave2 <- tm_basemap("Esri.WorldGrayCanvas") + tm_shape(merged) + tm_polygons(fill = "max_day_2") + tm_title("Date of Max Incidence (2nd Wave)") +   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )


cor9_total_cases_wave1 <- tm_basemap("Esri.WorldGrayCanvas") + tm_shape(merged) + tm_polygons(fill = "max_day_1") + tm_title("Number of Infections (1st Wave)") +   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )

cor10_total_cases_wave2 <- tm_basemap("Esri.WorldGrayCanvas") + tm_shape(merged) + tm_polygons(fill = "max_day_2") + tm_title("Number of Infections (2nd Wave)") +   tm_layout(
    legend.position = c("left", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,  
    legend.text.size = 1.2,
  legend.title.size = 1.4
  )

tmap_save(cor1_pop, "figures/cor1_pop.png")
tmap_save(cor2_pop_density, "figures/cor2_pop_density.png")
tmap_save(cor3_commuters, "figures/cor3_commuters.png")
tmap_save(cor4_commuter_density, "figures/cor4_commuter_density.png")
tmap_save(cor5_max_incidence_wave1, "figures/cor5_max_incidence_wave1.png")
tmap_save(cor6_max_incidence_wave2, "figures/cor6_max_incidence_wave2.png")
tmap_save(cor7_max_date_wave1, "figures/cor7_max_date_wave1.png")
tmap_save(cor8_max_date_wave2, "figures/cor8_max_date_wave2.png")
tmap_save(cor9_total_cases_wave1, "figures/cor9_total_cases_wave1.png")
tmap_save(cor10_total_cases_wave2, "figures/cor10_total_cases_wave2.png")
```


```{r}
library(dplyr)
library(tidyr)
library(purrr)

# vars_x <- c("commuters_total", "commuter_relative", "pop", "pop_density")
vars_x <- c("commuter_relative", "pop_density")


vars_y <- c(
  "max_incidence_1","max_day_1","total_cases_1",
  "max_incidence_2","max_day_2","total_cases_2"
)

pairs <- expand_grid(var1 = vars_x, var2 = vars_y)

result <- pairs %>%
  mutate(
    test = map2(var1, var2, ~ cor.test(merged[[.x]], merged[[.y]])),
    correlation = map_dbl(test, "estimate"),
    p_value = map_dbl(test, "p.value"),
    stars = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    corr_rounded = sprintf("%.3f%s", round(correlation, 3), stars)
  ) %>%
  select(var1, var2, correlation, p_value, corr_rounded)

result
```

# DEPRECATED
```{r}
### MAX WAVE HEIGHT
# 1st Wave - max wave height 
# no correlation Commuters -> Incidence:  -0.06574884 ->  -0.0259287
cor(merged$commuters_total,merged$max_incidence_1)

# no correlation Relative Commuters -> Incidence:  0.03138117 -> 0.0349035
cor(merged$commuter_relative,merged$max_incidence_1)

# a bit correlation Population -> Incidence: -0.2914974 -> no change
cor(merged$pop,merged$max_incidence_1)

# positive correlation 0.5235667 -> no change
cor(merged$pop_density,merged$max_incidence_1)


# 2nd Wave
# no correlation Commuters -> Incidence:  -0.01679962
cor(merged$commuters_total,merged$max_incidence_2)

# no correlation Relative Commuters -> Incidence:  0.02488187
cor(merged$commuter_relative,merged$max_incidence_2)

# a bit correlation Population -> Incidence: -0.2333074
cor(merged$pop,merged$max_incidence_2)

# positive correlation 0.3811721
cor(merged$pop_density,merged$max_incidence_2)


### Date of Max Wave
# 1st Wave 
# some correlation Commuters -> Date:  -0.4129234
# cor(merged$commuters_total,merged$max_day_1)

# some correlation Relative Commuters -> Date:  -0.424026
cor(merged$commuter_relative,merged$max_day_1)

# negative correlation Population -> Date:-0.4307691
# cor(merged$pop,merged$max_day_1)

# negative correlation -0.2520205
cor(merged$pop_density,merged$max_day_1)


# 2nd Wave 
# some correlation Commuters -> Date:  -0.0494883
cor(merged$commuters_total,merged$max_day_2)

# some correlation Relative Commuters -> Date:  -0.03293716
cor(merged$commuter_relative,merged$max_day_2)

# a bit correlation Population -> Date: 0.04996071
cor(merged$pop,merged$max_day_2)

# negative correlation: 0.1319583
cor(merged$pop_density,merged$max_day_2)


### Total Number of Infections
# 1st Wave 
# cor(merged$commuters_total,merged$total_cases_1)

# some correlation Relative Commuters
cor(merged$commuter_relative,merged$total_cases_1)

# negative correlation Population -> 
# cor(merged$pop,merged$total_cases_1)

# negative correlation
cor(merged$pop_density,merged$total_cases_1)


# 2nd Wave 
# some correlation Commuters -> 
# cor(merged$commuters_total,merged$total_cases_2)

# some correlation Relative Commuters ->
cor(merged$commuter_relative,merged$total_cases_2)

# a bit correlation Population ->
# cor(merged$pop,merged$total_cases_2)

# negative correlation:
cor(merged$pop_density,merged$total_cases_2)

# merged %>% view()
```


```{r}

merged %>% ggplot() + geom_point(aes(pop_density,total_cases_1))
merged %>% ggplot() + geom_point(aes(pop_density,max_incidence_1))
merged %>% ggplot() + geom_point(aes(pop_density,max_date_1))
```


```{r}
berlin <- read.delim("/Users/jakob/berlin.txt")
brand <- read.delim("/Users/jakob/brand.txt")
both <- read.delim("/Users/jakob/both.txt")

brand %>% filter(district == "total" & date == "2021-02-23") %>% pull(nShowingSymptomsCumulative)
both %>%  filter(district == "total" & date == "2021-02-23") %>% pull(nShowingSymptomsCumulative)
berlin %>% filter(district == "total" & date == "2021-02-23") %>% pull(nShowingSymptomsCumulative)
```

