---
title: "preprocess_RKI_data"
author: "Inan Bostanci"
date: "2025-07-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# import necessary libraries
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(zoo)
```

## RKI data on reported cases
RKI provides the daily number of new confirmed COVID-19 cases in Germany, available at the county level (IdLandkreis, starts with "11" for Berlin, "12" for Brandenburg), broken down by age group and gender. The data can be differentiated into "Meldedatum", which is the date of reporting, and "Refdatum", which is either the assumed date of the actual onset, or if unknown, the date of reporting. 

The data is available in CSV format and can be downloaded from https://github.com/robert-koch-institut/SARS-CoV-2-Infektionen_in_Deutschland_Archiv

For the ODE fit, it seems that we used "Refdatum", which is the date of the actual onset or reporting date, and not "Meldedatum", which is the date of reporting. However, for the ABM, I think we used "Meldedatum" - this might make sense if the ABM enables to simulate the reporting process, while the ODE's I-compartment simulates the actual disease progression.

```{r}
# Helper function to process and aggregate RKI data
process_rki_data <- function(df, date_col, region_prefix = "11") {
  df %>%
    filter(str_starts(as.character(IdLandkreis), region_prefix)) %>%
    group_by(date = as_date(.data[[date_col]])) %>%
    summarise(daily_new = sum(AnzahlFall, na.rm = TRUE), .groups = "drop") %>%
    complete(date = seq(min(date), max(date), by = "1 day"), fill = list(daily_new = 0)) %>%
    arrange(date) %>%
    mutate(
      active_cases = daily_new + lead(daily_new, 1, default = 0) + lead(daily_new, 2, default = 0),
      cumulative_cases = cumsum(daily_new),
      recovered = lag(cumulative_cases, default = 0),
      susceptible = 3782202 - cumulative_cases
    )
}
```

Load data from Github
```{r}
# Load data from the RKI GitHub repository
rki_repo_url <- "https://github.com/robert-koch-institut/SARS-CoV-2-Infektionen_in_Deutschland_Archiv/raw/refs/heads/main/Archiv/2021-05-31_Deutschland_SarsCov2_Infektionen.csv.xz"

# Download to a temp file
tmpfile <- tempfile(fileext = ".csv.xz")
download.file(rki_repo_url, tmpfile, mode = "wb")

# Read from compressed file directly
raw_data <- read_csv(tmpfile)
```


```{r}
# Load data and process both Refdatum and Meldedatum
# raw_data <- read_csv('/Users/inanb/Documents/GitHub/berlin_ode/2021-05-31_Deutschland_SarsCov2_Infektionen.csv')
berlin_ref <- process_rki_data(raw_data, "Refdatum")
berlin_melde <- process_rki_data(raw_data, "Meldedatum")
```


```{r}
# First, gather the data to long format for plotting
plot_data <- berlin_ref %>%
  select(date, daily_new, active_cases) %>%
  pivot_longer(cols = c(daily_new, active_cases),
               names_to = "type", values_to = "count")

# Plot
ggplot(plot_data, aes(x = date, y = count, color = type)) +
  geom_line(size = 1) +
  labs(title = "Daily New Cases and Active Cases in Berlin",
       x = "Date",
       y = "Number of Cases",
       color = "Type") +
  theme_minimal()

```


Lets compare the curves for Meldedatum and Refdatum

```{r}
# merge be_ref and be_melde for comparison
berlin_combined <- berlin_ref %>%
  select(date, daily_new_ref = daily_new, active_cases_ref = active_cases) %>%
  left_join(
    berlin_melde %>%
      select(date, daily_new_melde = daily_new, active_cases_melde = active_cases),
    by = "date"
  )

period_filter <- between(berlin_combined$date, as.Date("2020-03-02"), as.Date("2020-12-31"))

ggplot(filter(berlin_combined, period_filter), aes(x = date)) +
  geom_line(aes(y = daily_new_ref, color = "Refdatum")) +
  geom_line(aes(y = daily_new_melde, color = "Meldedatum")) +
  labs(title = "Comparison of Daily New Cases", x = "Date", y = "Daily Cases", color = "Date Type") +
  theme_minimal()


```
```{r}
# smooth the curves with rolling 7 day incidence per 100.000


tmpfile <- tempfile(fileext = ".csv")
download.file("https://raw.githubusercontent.com/robert-koch-institut/COVID-19_7-Tage-Inzidenz_in_Deutschland/refs/heads/main/COVID-19-Faelle_7-Tage-Inzidenz_Bundeslaender.csv", tmpfile, mode = "wb")

# Read from compressed file direct
xxx  <- read_csv(tmpfile)

xxx2 <- xxx %>% filter(Bundesland_id == "11") %>% filter(Altersgruppe == "00+")

incidence_data_be <- berlin_combined %>%
  mutate(
    incidence_ref = rollsum(daily_new_ref, 7, fill = NA, align = "right") / 3782202 * 1e5,
    incidence_melde = rollsum(daily_new_melde, 7, fill = NA, align = "right") / 3782202 * 1e5
  )

ggplot(filter(incidence_data_be, period_filter), aes(x = date)) +
  geom_line(aes(y = incidence_ref, color = "Refdatum")) +
  geom_line(aes(y = incidence_melde, color = "Meldedatum")) +
  geom_line(data = xxx2 %>% filter(Meldedatum < ymd("2020-12-31")), aes(Meldedatum,`Inzidenz_7-Tage`)) +
  labs(title = "7-Day Incidence per 100k (Berlin)", x = "Date", y = "Incidence", color = "Date Type") +
  theme_minimal()

```
The curve for Meldedatum has a delayed peak, while the curve for Refdatum has a peak around mid-March 2020. The Refdatum curve is more representative of the actual disease progression, while the Meldedatum curve reflects the reporting process, which can introduce delays. 

```{r}

# Normalize and build final berlin_data frame
berlin_data <- berlin_ref %>%
  mutate(susceptible = susceptible / 3782202,
         infected = active_cases / 3782202,
         recovered = recovered / 3782202) %>%
  mutate(exposed = 0,
         symptomatic = 0,
         hospitalized = 0,
         critical = 0,
         deceased = 0,
         quarantine_S = 0,
         quarantine_E = 0,
         quarantine_I = 0,
         contact_traced = 0) %>%
  select(date, susceptible, exposed, infected, recovered, symptomatic,
         hospitalized, critical, deceased, quarantine_S, quarantine_E,
         quarantine_I, contact_traced) %>%
  arrange(date)
# Save to CSV
# write_csv(berlin_data, "/Users/inanb/Documents/ZIB/ode_berlin_2502/data.csv")
```

Same for Brandenburg

```{r}
brandenburg_ref <- process_rki_data(raw_data, "Refdatum", region_prefix = "12")
brandenburg_melde <- process_rki_data(raw_data, "Meldedatum", region_prefix = "12")

# First, gather the data to long format for plotting
plot_data <- brandenburg_ref %>%
  select(date, daily_new, active_cases) %>%
  pivot_longer(cols = c(daily_new, active_cases),
               names_to = "type", values_to = "count")

# Plot
ggplot(plot_data, aes(x = date, y = count, color = type)) +
  geom_line(size = 1) +
  labs(title = "Daily New Cases and Active Cases in Brandenburg",
       x = "Date",
       y = "Number of Cases",
       color = "Type") +
  theme_minimal()
```

```{r}
# merge be_ref and be_melde for comparison
brandenburg_combined <-brandenburg_ref %>%
  select(date, daily_new_ref = daily_new, active_cases_ref = active_cases) %>%
  left_join(
    brandenburg_melde %>%
      select(date, daily_new_melde = daily_new, active_cases_melde = active_cases),
    by = "date"
  )

period_filter <- between(brandenburg_combined$date, as.Date("2020-03-02"), as.Date("2020-05-31"))

ggplot(filter(brandenburg_combined, period_filter), aes(x = date)) +
  geom_line(aes(y = daily_new_ref, color = "Refdatum")) +
  geom_line(aes(y = daily_new_melde, color = "Meldedatum")) +
  labs(title = "Comparison of Daily New Cases (Brandenburg)", x = "Date", y = "Daily Cases", color = "Date Type") +
  theme_minimal()

```

```{r}
# smooth the curves with rolling 7 day incidence per 100.000
incidence_data_br <- brandenburg_combined %>%
  mutate(
    incidence_ref = rollmean(daily_new_ref, 7, fill = NA, align = "right") / 2556747 * 1e5,
    incidence_melde = rollmean(daily_new_melde, 7, fill = NA, align = "right") / 2556747 * 1e5
  )

ggplot(filter(incidence_data_br, period_filter), aes(x = date)) +
  geom_line(aes(y = incidence_ref, color = "Refdatum")) +
  geom_line(aes(y = incidence_melde, color = "Meldedatum")) +
  labs(title = "7-Day Incidence per 100k (Brandenburg)", x = "Date", y = "Incidence", color = "Date Type") +
  theme_minimal()

```

Now, combined incidence in Berlin and Brandenburg
```{r}
bebr_combined <- berlin_combined %>%
  select(date, daily_new_ref, daily_new_melde) %>%
  rename(daily_new_berlin_ref = daily_new_ref, daily_new_berlin_melde = daily_new_melde) %>%
  left_join(
    brandenburg_combined %>%
      select(date, daily_new_ref, daily_new_melde) %>%
      rename(daily_new_brandenburg_ref = daily_new_ref, daily_new_brandenburg_melde = daily_new_melde),
    by = "date"
  )

# sum the daily new cases for both regions
bebr_combined <- bebr_combined %>%
  mutate(
    daily_new_ref = daily_new_berlin_ref + daily_new_brandenburg_ref,
    daily_new_melde = daily_new_berlin_melde + daily_new_brandenburg_melde
  )

period_filter <- between(bebr_combined$date, as.Date("2020-03-02"), as.Date("2020-05-31"))

ggplot(filter(bebr_combined, period_filter), aes(x = date)) +
  geom_line(aes(y = daily_new_ref, color = "Refdatum")) +
  geom_line(aes(y = daily_new_melde, color = "Meldedatum")) +
  labs(title = "Comparison of Daily New Cases (Berlin + Brandenburg)", x = "Date", y = "Daily Cases", color = "Date Type") +
  theme_minimal()

```
Incidence
```{r}
# smooth the curves with rolling 7 day incidence per 100.000
incidence_data_bebr <- bebr_combined %>%
  mutate(
    incidence_ref = rollmean(daily_new_ref, 7, fill = NA, align = "right") / (2556747 + 3782202) * 1e5,
    incidence_melde = rollmean(daily_new_melde, 7, fill = NA, align = "right") / (2556747 + 3782202) * 1e5
  )

ggplot(filter(incidence_data_bebr, period_filter), aes(x = date)) +
  geom_line(aes(y = incidence_ref, color = "Refdatum")) +
  geom_line(aes(y = incidence_melde, color = "Meldedatum")) +
  labs(title = "7-Day Incidence per 100k (Berlin + Brandenburg)", x = "Date", y = "Incidence", color = "Date Type") +
  theme_minimal()
```

Plot the incidence for the full year
```{r}
period_filter_2020 <- between(bebr_combined$date, as.Date("2020-03-02"), as.Date("2020-12-31"))
ggplot(filter(incidence_data_bebr, period_filter_2020), aes(x = date)) +
  geom_line(aes(y = incidence_ref, color = "Refdatum")) +
  geom_line(aes(y = incidence_melde, color = "Meldedatum")) +
  labs(title = "7-Day Incidence per 100k (Berlin + Brandenburg)", x = "Date", y = "Incidence", color = "Date Type") +
  theme_minimal()
```

